Cache Oblivious B-trees

### 2011-12-09 paper.tex
Adding the first pass at the background section. It is a piece of crap right now.

Bender et al. first introduced a cache oblivious B-tree (CO B-tree) in
\cite{cobtree}. The CO B-tree was able to match the optimum search bound of
$\Theta(log_B N)$ memory transfers without including any specific
parameterization describing the speed or sizes of the machine's memory
hierarchy.

The CO B-tree paper presents three different data structures that make
different trade off in the number of memory transfer they perform. However,
all of them achieve their performance by structuring their elements as a
strongly weight-balanced search tree (SWBST). The SWBST is then laid out into
a packed memory array (PMA) using a van Emde Boas (vEB) layout. The SWBST and
PMA play crucial roles in the other cache-oblivious data structures that were
developed after the CO B-tree and thus are described at length in the next two
sections.

@@

Strongly weight-balanced search trees play a key role in CO B-tree. The weight
of a node is $u$ is equal to the total number of $u$'s descendants plus 1. It
can als be thought of as the sum of the weights of $u$'s children plus 1.
Formally, $u$'s weight is: $w(u) = 1 + \sum_{v \in children} w(v)$.

Most weight-balanced search trees which requires that the heights of a node's
left and right subtrees are simply off by a constant factor. SWBST enforce the
stronger property that a node at height $h$ in the tree (height is the node's
distance from the leaves; leaves have a height of 1) has $\Theta c^h$ where
$c$ is some constant.

A known data structure that achieves this property is the ``weight-balanced
B-tree'' \cite{paper}. The CO B-tree leverages this data structure to maintain
its elements. The next section describes how SWBST perform various operations
while maintain their strongly weight-balanced property.

\subsection{Operations}

The CO B-tree supports \Search, \Insert, \Delete, and \Scan operations.

It should be noted, that keeping the level of indirection sorted is only
needed in case we want to perform \Scan operations. If we do not need \Scan,
then we can keep the indirection array in any arbitrary order reducing the
cost of \Insert and \Delete by the $(log^2 N) /B$ factor making them incur
$O(log_{B} N)$ memory transfers overall.

SWBST is only one half of the CO B-tree story...

@@

>>>>

Bender et al. first introduced a cache oblivious B-tree (CO B-tree) in
\cite{cobtree}. The paper presents three different data structures that make
different trade offs in the number of memory transfer they perform. However,
all of them achieve their performance by structuring their elements as a
strongly weight-balanced search tree (SWBST). The SWBST is then laid out into
a packed memory array (PMA) using a van Emde Boas (vEB) layout. The SWBST and
PMA play crucial roles in the other cache-oblivious data structures that were
developed after the CO B-tree and thus are described at length in the next two
sections.

@@

@@

The ultimate goal of a CO B-tree is to reduce the number of memory transfers
while supporting \Search{}, \Insert{}, and \Delete{} operations. Thus, at a
fundamental level the CO B-tree still needs to maintain its data in some
structure. That structure is a balanced search tree. The search tree is laid
out according to a van Emde Boas (vEB) layout [???, ???]. The idea of a vEB
layout is fairly straight forward. Let $h$ denote the height of a tree. the
vEB layout splits the tree approximately at height $h/2$. This splits the
original tree into a ``top'' subtree and many ``bottom'' subtrees. Let $A$
denote the top subtree and $B_1... B_k$ denote the bottom subtrees. The vEB
layout is defined recursively using the order: $A, B1, ... B_k$. This layout
is also shown in Figure~\ref{fig:veb}.

A nice consequence of using a vEB layout is that we know that it will use
$O(log_B N)$ memory transfers for searches [???, ???]. Thus, off the bat we
have achieve the desired number of memory transfers for search queries. It
still remains to be shown that the CO B-tree achieves the desired number of
memory transfers for dynamic inserts and deletes.

We know that the CO B-tree is going to use a balance search tree, but what
kind?. Strongly weight-balanced search trees play a key role in CO B-tree. The
weight of a node is $u$ is equal to the total number of $u$'s descendants plus
1. It can also be thought of as the sum of the weights of $u$'s children plus
1. Formally, $u$'s weight is: $w(u) = 1 + \sum_{v \in children} w(v)$.

Most weight-balanced search trees which requires that the heights of a node's
left and right subtrees are simply off by a constant factor. SWBST enforce the
stronger property that a node at height $h$ in the tree (height is the node's
distance from the leaves; leaves have a height of 1) has $\Theta c^h$
descendants where $c$ is some constant. This property becomes important later
on when laying the tree out in memory (see Section~\ref{}).

A known data structure that achieves this property is the ``weight-balanced
B-tree'' \cite{????}. The CO B-tree leverages this data structure to maintain
its elements. Before SWBST can be used, however, it remains to be proven that
it maintains the the $c^h$ descendants property even after inserts and
deletes.

The definition of a SWBST given in [????] provides us with this bound for the
weight of a node:

\begin{equation}
\frac{c^{h-1}}{2} \leq w(u) \leq 2c^{h-1}}
\end{equation}

Lets now consider insertions. To insert an element into a WSBST we search down
the tree and find the natural place to insert the element - just like in any
tree structure. After inserting the element into some node $w$, we have to
check to see if any of $w$'s ancestors have become unbalanced by having a
weight greater than $2^{d-1}$. If such an ancestor exists, it follows that
every node along the path between that ancestor and $w$ will be un balanced as
well. Thus, we start with $w$ and move up the tree.

When we split a node, we naturally need to determine which children belong to
which of the two nodes. We cannot just divide the children evenly because it
may cause another violation. Thus, instead, we find the longest sequence of
elements such that the weight of that subsequence is $\lceil w(u)/2 \rceil$.
By doing this, we maintain the bound that $\frac{c^{h-1}{2} \leq w(u) \leq
2c^{h-1}}$. We cascade these changes up the tree until all of the ancestors
are balanced.

Deletions are supported in the exact same way, except, we check is a node has
a weight $w(u) < \frac{c^{h-1}}{2}$. In this case, we will merge two neighbors
together and proceed up the tree. There is one edge case that needs to be
handled. After merging, you may produce a new node that has a weight larger
than the bounds. If this happens we immediately split the node again. Thus,
this can be thought of as stealing nodes from a neighbor.

WE have now shown that a SWBST achieves the desired property that the number
of descendants of a node is $c^h$. We can now show how to lay this tree out in
memory.

Now that we have our elements structured in a SWBST, we are still left with
mapping this tree to memory and proving that it achieves our desired number of
memory transfers. The packed memory array (PMA) was developed to solve this
problem [???]. There are three main costs that the PMA is concerned with.

Fist, the PMA needs to maintain open space in its memory for newly inserted
items. The PMA is able to accomplish this using $O(1 + (log^2 N) / B)$ memory
transfers. Doing so may seem like attempting to achieve contradictory goals:
maintaining the locality of elements while keeping space available for new
inserts. The PMA uses a technique described in [???]. At a high level, the PMA
maintains a windows into the array. When a window becomes too full or too
empty, the elements are spread out within the context of a larger window. Two
parameters are required, the window size and the threshold for trigger the
spreading out. However, these parameters are not related to the block size of
our memory level and thus do not violate the CO model of computation. Overall,
the PMA is able to keep extra space in $O(1 + (log^2 N) / B)$ memory
transfers.

Second, the PMA needs to handle the splitting and merging that will be caused
by the SWBST discussed in the previous section. THe PMA is able to accomplish
this using $O(1 + (log N) /B)$ memory transfers. The proof is based on the
fact that each node has $c^h$ descendants (recall, this is the key property
that SWBST provided). Thus, merging or splitting a node will cause $c^h$ nodes
to move, requiring $O(1 + c^h/B)$ memory transfers. It follows from the
defitinot of SWBST in [???] that when a node is rebalanced, it will require
$c^h$ insertions or deletions until it is rebalanced again. Thus, the
amortized cost of each insertion and/or deletion is $O(1/B)$. When you split a
node, it may go all the way up the tree and thus overall it will incur $O((log
N) / B)$ memory transfers.

Third, the PMA needs to handle updating the parent and child pointers and may
become invalidated because of how the nodes moved around. The PMA is able to
update the parent and child pointers. If the pointers are all close together
in memory (e.g. within $B$ distance of each other) it is not big deal because
only a constant number of memory transfers are required. However, if they are
further apart the PMA introduces buffer nodes. The buffer nodes don't actually
do anything. All they do is serve to ``stretch'' the vEB layout so that nodes
do not move far away from one another. By using these buffers the PMA is able
to achieve a the pointer update cost using $O(1+ logB / \sqrt{B}) log^2 N)$
memory transfers.

Of these three terms, the dominating factor is maintain room for newly
inserted elements - $O((log^2N)/B)$. We know that SWBST achieve insertions and
deletitions in $O(log_B N)$ if they are cache-aware. Thus, adding PMA makes
the cache-oblivious and thereby incurring an additional $O((log^2N)/B)$ memory
transfers.

Therefore, overall, CO B-Tree achieve insertions and deletions in $O(log_B N +
(log^2N)/B)$ memory transfers. However, they still achieve the optimal
$O(log_B N)$ memory transfers for searches, matching the performance of
ordinary, cache-aware, B-trees.
